Real-space computation of $E$/$B$-mode maps I: Formalism and Compact Kernels

### 2017-12-05 articles/formalism/sections/real_space_formalism.tex
Minor edits to Euler angles plot/movie. Added appendix in formalism paper deriving the important identity on product of spin spherical harmonics and the radiating and convolution interpretations of the kernel.

\caption{}

@@

\textit{Euler angles:}  We define the local cartesian coordinate system at any point on the sphere such that the z-axis points along the radial direction, the x-axis is along the vector tangent to the local longitude pointing south and the y-axis is a vector tangent to the local latitude pointing east. The Euler angles $\alpha ,\beta ~\&~ \gamma$ define rotation operations that transforms the local cartesian coordinate system defined at the location $\hat{n}_0 \equiv (\theta_0,\phi_0)$ such that it aligns with the local cartesian coordinate system at the location $\hat{n}_i \equiv (\theta_i,\phi_i)$ \cite{varshalovich}.

Specifically $\alpha$ defines the rotation about the z-axis, $\beta$ defines the rotation about the new y-axis (y1-axis) after the previous rotation and $\gamma$ defines the rotation about the final z-axis (z2-axis) after carrying out the previous two rotations. In more physical terms, these angles can be understood as follow: The rotation by $\alpha$ about the z-axis is such that it aligns the x-axis of the cartesian system at location $\hat{n}_0$ along the great circle in the direction of $\hat{n}_i$.  The rotation by $\beta$ about the y2-axis parallel transports the local cartesian coordinate system from location $\hat{n}_0$ to location $\hat{n}_i$, such that the z-axes of the two coordinate systems are parallel to each other. Finally the rotation by angle $\gamma$ about the z2-axis aligns the x \& y axes of the parallel transported system with those of the local cartesian system defined at $\hat{n}_i$.

The simplest case is when one of the coordinates coincides with the north pole (more formally this refers to the point $\theta_0 \rightarrow 0$ while moving along the longitude $\phi_0=0$), say $\hat{n}_0=(\theta_0,\phi_0)=(0,0)$. In this case it is easy to visualize and infer that the rotations by Euler angles: $(\alpha,\beta,\gamma) =(\phi_i,\theta_i,0)$ in the $z-y1-z2$ sense, will align the local cartesian system at the pole with that defined at position $\hat{n}_i$. The Euler angles in the $z-y-z$ convention are related to those in the $z-y1-z2$ convection by the following rule: $(\alpha,\beta,\gamma)_{z-y-z} =(\gamma,\beta,\alpha)_{z-y1-z2}$.

@@

where the symbol ${}_{0}Y_e$ is used to denote the sub-matrix ${}_{0}Y_{\hat{n}_e \times \ell m} \equiv {}_{0}Y_{\ell m}(\hat{n}_e)$, the symbol ${}_{\pm 2}Y^{T*}_q$ is used to denote the matrix ${}_{\pm 2}Y^*_{\ell m \times \hat{n}_q} \equiv {}_{\pm 2}Y^*_{\ell m}(\hat{n}_q)$ and the summation is over the multipole indices $\ell,m$. \revisit{We purposefully use the index ``e'' to denote the location where the scalar fields are measured and the index ``q'' to denote the location where the Stokes parameters are measured.} Using the conjugation properties of the spin spherical harmonic functions it can be shown that the following relation holds true, \comment{These constructions are dyadics}

@@

 where the terms on either side of the equation are those that appear in \eq{eq:qu2eb_ker_1}. The $m$ sum over the product of two spherical harmonic functions with spins $s_1$ and $s_2$ respectively, is given by the following identity \cite{varshalovich},

@@

\caption{}

>>>>

\caption{This figure depicts the Euler angles in the z-y1-z2 convention. The cartesian coordinates shown in dark green are those that lie in the tangent plane at location $\hat{n}_q = (\theta_q, \phi_q)$ while those shown in dark red are the ones that lie in the tangent plane at location $\hat{n}_e = (\theta_e, \phi_e)$. The blue coordinates at different locations are representative of the parallel transport along the geodesic connection the two locations $\hat{n}_q$ \& $\hat{n}_e$ on the sphere.}

@@

\textit{Euler angles:}  We define the local cartesian coordinate system at any point on the sphere such that the z-axis points along the radial direction, the x-axis is along the vector tangent to the local longitude pointing south and the y-axis is a vector tangent to the local latitude pointing east as shown in \fig{fig:euler_angles}. The Euler angles $\alpha ,\beta ~\&~ \gamma$ define rotation operations that transforms the local cartesian coordinate system defined at the location $\hat{n}_0 \equiv (\theta_0,\phi_0)$ such that it aligns with the local cartesian coordinate system at the location $\hat{n}_i \equiv (\theta_i,\phi_i)$ \cite{varshalovich}.

For this work, we believe it is convenient to work in the z-y1-z2 convention, therefore all future references to Euler angles are to be assumed to be given in this convention. \footnote{The Euler angles in the more standard $z-y-z$ convention are related to those in the $z-y1-z2$ convection by the following rule: $(\alpha,\beta,\gamma)_{z-y-z} =(\gamma,\beta,\alpha)_{z-y1-z2}$ \cite{varshalovich}.}
In the z-y1-z2 convention, $\alpha$ defines the rotation about the z-axis, $\beta$ defines the rotation about the new y-axis (y1-axis) after the previous rotation and $\gamma$ defines the rotation about the final z-axis (z2-axis) after carrying out the previous two rotations. In more \revisit{physical terms}, these angles can be understood as follows: The rotation by $\alpha$ about the z-axis is such that it aligns the x-axis of the cartesian system at location $\hat{n}_0$ along the great circle in the direction of $\hat{n}_i$.  The rotation by $\beta$ about the y2-axis parallel transports the local cartesian coordinate system from location $\hat{n}_0$ to location $\hat{n}_i$, such that the z-axes of the two coordinate systems are parallel to each other. Finally the rotation by angle $\gamma$ about the z2-axis aligns the x \& y axes of the parallel transported system with those of the local cartesian system defined at $\hat{n}_i$. The different Euler angles and the corresponding rotations are depicted in \fig{fig:euler_angles}.

The simplest scenario is when one of the coordinates coincides with the north pole $\hat{n}_0=(\theta_0,\phi_0)=(0,0)$ (the local cartesian system at this point being defined with an arbitrary convention set by choosing $\theta_0 \rightarrow 0$ while moving along the longitude $\phi_0=0$), in which case it is easy to see that rotations by Euler angles: $(\alpha,\beta,\gamma) =(\phi_i,\theta_i,0)$ will align the cartesian system at $\hat{n}_0$ with that defined at position $\hat{n}_i$. 

@@

where the symbol ${}_{0}Y_e$ is used to denote the sub-matrix ${}_{0}Y_{\hat{n}_e \times \ell m} \equiv {}_{0}Y_{\ell m}(\hat{n}_e)$, the symbol ${}_{\pm 2}Y^{T*}_q$ is used to denote the matrix ${}_{\pm 2}Y^*_{\ell m \times \hat{n}_q} \equiv {}_{\pm 2}Y^*_{\ell m}(\hat{n}_q)$ and the summation is over the multipole indices $\ell,m$. \revisit{We purposefully use the index ``e'' to denote the location where the scalar fields are being evaluated and the index ``q'' to denote the location from which  the Stokes parameters are being accessed.} Using the conjugation properties of the spin spherical harmonic functions it can be shown that the following relation holds true, \comment{These constructions are dyadics}

@@

 where the terms on either side of the equation are those that appear in \eq{eq:qu2eb_ker_1}. Note that each sub-matrix in \eq{eq:qu2eb_ker_1} is formed by some linear combination of a complex number and its conjugate. The $m$ sum over the product of two spherical harmonic functions with spins $s_1$ and $s_2$ respectively, is given by the following identity \cite{varshalovich},

@@

\caption{The local cartesian coordinate are only drawn on the red circle, representative of the coordinate dependence of the Stokes parameters. The coordinates are not drawn on the blue circle representing the fact that the value of the scalar field does not depend of the coordinate definition. The value of the scalar field at the specific location `e' can be evaluated by summing over the contribution from all the Stokes parameters on the red circle (sphere). The convolutions are performed with kernels which are defined in term of the Euler angle $\gamma_{eq}$. Alternatively, if centered on `q' then one can compute the contribution from the Stokes parameter at this specific location to all the point on the blue circle(sphere). The contribution across the sky from the Stokes parameter at the location `q'  is given by a function of the Euler angle $\alpha_{eq}$. In the flat sky limit, since $\gamma=-\alpha$ there is no difference between the radiating and convolving kernels.}
